### 2020-03-04 doc/icfp20/harmony.tex
major work on harmony section

@@

Harmony is a particularly fundamental area of music theory with numerous texts, for
example \citet{piston-harmony} and \citet{aldwell2018harmony}, devoted
to it. It has also been the focus of most of the Haskell-based
work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp}, and as a
first step into this huge area it worth trying to implement some of
this functionality in Agda. Indeed \citet{magalhaes-harmtrace} states in its final section
``It would be interesting to see if we could easily port our system to
a dependently-typed setting'' and explicitly mentions Agda and Idris
as candidates. Our work, while currently very preliminary, provides a
glimpse into what that can look like.

\citet{magalhaes-harmtrace} aim to creating a grammar for
harmonic analysis and a corresponding parser that takes as input a
series of chords and tries to produce a harmonic analysis that fits
the progression. The grammar makes crucial use of dependent types and
is thus awkward to represent in Haskell. We have implemented those
parts in Agda (\texttt{Fmmh.agda}) and not surprisingly they are simple and
elegant. However \citet{magalhaes-harmtrace} also makes use of
sophisticated Haskell libraries for parser combinators and generics,
which are unavailable in Agda and would either have to be natively
implemented or called through Agda's Haskell FFI, both of which involve
considerable work, so we have not yet pursued that route.

Instead we first focus on \fharm~\citep{koops-fharm}, which aims to create a
harmonize a given melody. This is the subject of Chapter~9 of 
\citet{piston-harmony}, and, along with three exercises,
\fharm harmonizes the main Example~9.1 of the chapter,
shown in Figure~\ref{fig:harmonization}.

@@

@@

Not having an implementation of the grammar and parser we explore a
different approach, based closer on Piston's.

TODO: More here.

>>>>

\newcommand{\HarmonizationMT}{
  \begin{figure}
    \includegraphics[width=5.5cm]{fig/piston.png}
    \hspace{1cm}
    \includegraphics[width=5cm]{fig/fharm.png} \\
    \begin{flushleft}
    \begin{small}
     \hspace{2.35cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \ \ \,IV \,VI \ \,III \ IV \ \ \,V
     \hspace{2.6cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \,IV VI \,III\,IV \ \ V
    \end{small}
    \end{flushleft}
    \caption{Harmonization by Music Tools (to be replaced by real figure)}
    \Description{Harmonization by Music Tools}
    \label{fig:harmonizationMT}
  \end{figure}
}

@@

Harmony refers to simultaneously sounding tones (chords) and its study
is primarily concerned with the progression of chords throughout a
piece of music. It is in essence the of dual of counterpoint. In this
section we formalize a small part of the classic text
\citet{piston-harmony}, in particular part of Chapter~9 which is
concerned with techniques for harmonizing a given melody.

\subsection{Harmonic Progression}
\label{sec:harmony:prog}

The primary Example~9.1 in Piston's text is shown on the left
side of Figure~\ref{fig:harmonization} which shows the melody with a
sample four part SATB harmonization in C~major. The melody is the
series of highest notes, the soprano (S) line, while the other four
parts alto (A), tenor (T) and bass (B) are the melodies below
it. Roman numerals denote the names of the triads at each vertical
slice. Note that even though there are four voices there are only
three distinct pitches sounding at each vertical slice (where tones an
octave apart are considered to have the same relative pitch). As is
typical the root of the triad is always doubled. Also note that all
triads are in root position (meaning the root is in the bass voice),
as Piston restricts to this case initially in the chapter (he does
relax it later). The goal of a harmonization is both to create an
aesthetically pleasing chord progression as well as fluid melodies in
each voice (called voice leading) and independence of each music
line (in other words good counterpoint). In this chapter as well as
throughout the text Piston gives advice on how to achieve these goals.

@@

We would like to formalize some of this advice and apply it to
automate harmonization of melodies. We start with selection of the
sequence of triads, in other words the harmonic progression. Earlier
in the text on page~23 Piston presents a Table of Usual Root
Prrogressions, noting which triads usually follow a given one and
which are sometimes used or rare. It is easy to encode this in Agda.

\begin{alltt}
data Triad : Set where
  I : Triad; II : Triad; III : Triad; IV : Triad; V : Triad; VI : Triad; VII : Triad

record NextTriad : Set where
  constructor nextTriadb
  field
    usual : List Triad; sometimes : List Triad; rare : List Triad

rootProgression : Triad \(\rightarrow\) NextTriad
rootProgression I   = nextTriad (IV :: V :: []) (VI :: []) (II :: III :: [])
rootProgression II  = nextTriad (V :: []) (IV  :: VI :: []) (I :: III :: [])
-- other cases omitted for space
\end{alltt}

It is easiest to create a harmonic progression in reverse, to
ensure one ends a musical phrase with either a full (V--I) or half (V)
cadence, and work backward from there. So we would like to invert this
table to give a list of likely triads to precede a given triad. This
would normally be straightfoward but here we see a disadvantage of
using Agda. As is typical with languages based on dependent type
theories, equality is a very subtle concept and even determing if a
term is a member of a list is nontrivial.

In Haskell this would be easy--there is an \texttt{Eq} typeclass and a
method for automatically generating equalities. In the Agda Standard
Library~\citep{agda-stdlib} is a relation with constructive proof of
membership, and requires far more effort than we need. Alternatives
include defining decidable equality for triads (which requires writing
out 49 cases) or mapping to natural numbers and using already
implemented decidable or boolean equality for those. For now we simply
write the inverse function manually, excluding the rare cases.

\begin{alltt}
previousTriads : Triad → List Triad
previousTriads I   = V :: IV :: VII :: []
previousTriads II  = VI :: IV :: []
-- other cases omitted for space
\end{alltt}

\subsection{Completing the Harmonization}
\label{sec:harmony:complete}

TODO: Finish this.

See Figure~\ref{fig:harmonizationMT}, which is currently just a copy of
the other figure, but included here now for space.

\HarmonizationMT

\subsection{Comparison with Previous Work}
\label{sec:harmony:compare}

Most of the the Haskell-based
work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp} has
focused on harmony. In particular \fharm~\citep{koops-fharm} tackles
the problem of harmonization of a melody, and their harmonization of
Piston's example is shown on the right side of
Figure~\ref{fig:harmonization}. Their method is quite different from
ours. Instead of using Piston's table of root progressions they use a
more sophisticated hierarchical grammar from previous
work~\citep{magalhaes-harmtrace}.

@@

TODO: More comparison.

### 2020-03-04 doc/icfp20/music.tex
major work on harmony section

A \emph{chord} refers to a set of simultaneously sounding notes.
In this paper we are particularly interested in \emph{triads},
which are chords consisting of three notes separated by the 3rd
intervals (i.e., \texttt{maj3} and \texttt{min3}).
We define triads as a datatype \texttt{Triad} with roman numeral
constructors representing the lowest note (called the \emph{root}).
For instance, \texttt{I} in C major represents the triad C-E-G.

>>>>

A \emph{chord} refers to a set of simultaneously sounding notes.  In
this paper we are particularly interested in \emph{triads}, which are
chords consisting of three notes separated by the 3rd intervals (i.e.,
\texttt{maj3} and \texttt{min3}).  We define triads as a datatype
\texttt{Triad} with roman numeral constructors representing the scale
degree of the lowest note (called the \emph{root}). The other two
notes are then members of the scale a third and a fifth above the
root. For instance, \texttt{I} in C major represents the triad C-E-G.
